---
title: "Predictions using the Human Activity Recognition Dataset"
author: "Csaba Iranyi"
output: 
  html_document
---

## The Question

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it.  
 
In this project, I will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants.  
Six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions:  
  
Movement type | Class
------------- | -------------
Exactly according to the specification  | Class A
Throwing the elbows to the front  | Class B
Lifting the dumbbell only halfway  | Class C
Lowering the dumbbell only halfway  | Class D
Throwing the hips to the fron  | Class E
  
Based on a dataset provide by Human Activity Recognition (HAR) [http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har](http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har) I will try:  

* to train and evaluate some multi-class classification predictive models using the training dataset with 159 features and one label (classe),
* to use the trained prediction models to predict 20 different test cases (what exercise was performed) from the testing dataset.

## Getting data

```{r warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width=10, fig.height=5)
options(width=120)

library(caret)
library(randomForest)
library(ggcorrplot)
library(ggpubr)
```

Downloading of traning and testing datasets.  

```{r}
# Downloading training dataset
training.data <- read.table(file = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", 
                            header = TRUE, 
                            sep = ",", 
                            encoding = "ISO_8859-1")

# Downloading training dataset
testing.data <- read.table(file = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", 
                           header = TRUE, 
                           sep = ",", 
                           encoding = "ISO_8859-1")
```

## Exploratory data analysis

Examining the dimensions and the features of the training dataset. 
```{r}
# Number of features
cat("Number of features in training dataset:", length(names(training.data)) - 1, "\n")
cat("Number of features in testing dataset:", length(names(training.data)) - 1, "\n")

# Number of rows
cat("Number of observations in training dataset:", nrow(training.data), "\n")
cat("Number of observations in testing dataset:", nrow(testing.data), "\n")

# Create summary of features (and label)
feature.summary <- data.frame(row.names = 1:length(names(training.data)))
feature.summary$feature.index <- 1:length(names(training.data))
feature.summary$feature.name <- names(training.data)
feature.summary$type <- sapply(training.data, class)
feature.summary$NA.rows <- sapply(training.data, function(x) sum(is.na(x)))
feature.summary$NA.rows.percent <- sapply(training.data, function(x) round(sum(is.na(x)) / length(x) * 100.0, digits = 2)) 
feature.summary$empty.rows <- sapply(training.data, function(x) length(x[x == ""]))
feature.summary$empty.rows.percent <- sapply(training.data, function(x) round(length(x[x == ""]) / length(x) * 100.0, digits = 2))
feature.summary$unique.values <- sapply(training.data, function(x) length(unique(x)))
feature.summary[, -1]
```

Lot of features has either NA or empty values in great percent.

## Cleaning data

### Removing useless features

Removing constant features and features with NA or empty rows from the training and testing data as well.
```{r}

# Removing useless features
#  - Constant features
#  - Features with NA rows
#  - Features with empty rows
useless.feature.indices <- feature.summary[feature.summary$unique.values == 1 | feature.summary$NA.rows != 0 | feature.summary$empty.rows !=0, 1]
cat("Number of removable useless features:", length(useless.feature.indices), "\n")
training.data <- training.data[, -useless.feature.indices]
testing.data <- testing.data[, -useless.feature.indices]
```

### Removing unnecessary features

Removing the sequence feature (X) and all time related features () from the training and testing data.
```{r}
# Removing unnecessary features
unnecessary.feature.indices <- grep("timestamp|X", names(training.data))
cat("Number of removable unnecessary features:", length(unnecessary.feature.indices), "\n")
training.data <- training.data[, -unnecessary.feature.indices]
testing.data <- testing.data[, -unnecessary.feature.indices]
```

### Converting  features

Converting all features to numeric type (except the label column).
```{r}
# Converting all features to numeric type (except the label column)
label.column <- training.data$classe
training.data <- data.frame(data.matrix(training.data))
training.data$classe <- label.column
testing.data <- data.frame(data.matrix(testing.data))

# Number of usable features
cat("Number of usable features:", length(names(training.data)) - 1, "\n")
```

## Splitting data

The downloaded test dataset (20 observations) is the the ultimate validation set (one time scoring).  
Splitting up the downloaded training dataset into a cross-validating (25%) and training dataset (75%).
```{r}
# Initialize RNG
set.seed(333)

# Crete training/CV datasets
partition.indices <- createDataPartition(y = training.data$classe, p = 0.75, list = FALSE)
cv.data <- training.data[-partition.indices,]
training.data <- training.data[partition.indices,]

cat("Number of rows in training dataset:", nrow(training.data), "\n")
cat("Number of rows in cross-validation dataset:", nrow(cv.data), "\n")
cat("Number of rows in testing dataset:", nrow(testing.data), "\n")
```

## Feature engineering

### Finding feature correlations

Calculating feature correlations with the outcome (classe).
```{r}
# Label (outcome) column index
label.index <- which(names(training.data) == "classe")

# Create summary of usable features
feature.set <- training.data[,-label.index]
feature.summary <- data.frame(row.names = 1:length(names(feature.set)))
feature.summary$feature.index <- 1:length(names(feature.set))
feature.summary$feature.name <- names(feature.set)
feature.summary$type <- sapply(feature.set, class)
feature.summary$unique <- sapply(feature.set, function(x) length(unique(x)))
feature.summary$mean <- sapply(feature.set, function(x) round(mean(x), digits = 4))
feature.summary$sum <- sapply(feature.set, function(x) round(sum(x), digits = 4))
feature.summary$sd <- sapply(feature.set, function(x) round(sd(x), digits = 4))
feature.summary$cor <- sapply(feature.set, function(x) round(abs(cor(x, as.numeric(training.data$classe))), digits = 4))
feature.summary <- feature.summary[order(feature.summary$cor, decreasing = TRUE),]
feature.summary[, -1]
```

Selecting 33 (manually choosen number) features with best correlations with outcome.
```{r}
# The most relevant features (best correlations with outcome)
most.correlated.outcome.feature.indices <- feature.summary$feature.index[1:33]
cat("Relevant features:", feature.summary$feature.name[1], "...", feature.summary$feature.name[33], "\n")
```

Searching for duplicated (lineary dependent) features to reduce pair-wise correlations.
```{r}
# Duplicated feature
feature.correlation.matrix <- cor(training.data[, -label.index])
duplicated.feature.indices <- findCorrelation(feature.correlation.matrix, cutoff = 0.9, exact = TRUE)
cat("Removable duplicated features:", paste(names(training.data)[duplicated.feature.indices], collapse = ", "), "\n")

# Plot correlation matrix
ggcorrplot(feature.correlation.matrix, hc.order = TRUE, type = "lower", colors = c("red", "white", "red"), insig = "blank", tl.cex = list(size = 4))
```
  
## Training and scoring models 

Creating a summary dataset to store result of trained models.
```{r}
# Summary of models
model.summary <- data.frame(row.names = 1:6, stringsAsFactors = TRUE)

# Initialize RNG
set.seed(333)
```

Setting options of random forest.
```{r}
# Number of trees
number.trees <- 50
```
 
### Random Forest: using all features 

```{r}
# Random Forest: all features
start.time <- Sys.time()
model.all.features <- randomForest(
  x = training.data[, -label.index], 
  y = training.data$classe,
  xtest = cv.data[, -label.index], 
  ytest = cv.data$classe, 
  ntree = number.trees,
  keep.forest = TRUE,
  proximity = TRUE,
  do.trace = FALSE)
end.time <- Sys.time()

model.summary <- rbind(model.summary, cbind( 
  trees = model.all.features$ntree,
  features = ncol(training.data) - 1,
  model = "All features",
  training.accuracy = round((1 - sum(model.all.features$confusion[,"class.error"])) * 100, digits = 3),
  cv.accuracy =round((1 - sum(model.all.features$test$confusion[,"class.error"])) * 100, digits = 3),
  training.time = round(as.numeric(end.time - start.time, units = "secs"))
))
```

### Random Forest: all features with PCA 

```{r}
# Apply PCA
pre.pca <- preProcess(training.data[, -label.index], method = "pca", thresh = 0.99)
training.data.pca <- predict(pre.pca, training.data[, -label.index])
cv.data.pca <- predict(pre.pca, cv.data[, -label.index])
testing.data.pca <- predict(pre.pca, testing.data[, -label.index])

# Random Forest: all feature with PCA
start.time <- Sys.time()
model.pca.features <- randomForest(
  x = training.data.pca, 
  y = training.data$classe,
  xtest = cv.data.pca, 
  ytest = cv.data$classe, 
  ntree = number.trees,
  keep.forest = TRUE,
  proximity = TRUE,
  do.trace = FALSE)
end.time <- Sys.time()

model.summary <- rbind(model.summary, cbind( 
  trees = model.pca.features$ntree,
  features = ncol(training.data.pca),
  model = "PCA",
  training.accuracy = round((1 - sum(model.pca.features$confusion[,"class.error"])) * 100, digits = 3),
  cv.accuracy = round((1 - sum(model.pca.features$test$confusion[,"class.error"])) * 100, digits = 3),
  training.time = round(as.numeric(end.time - start.time, units = "secs"))
))
```

### Random Forest: the most correlated features 

```{r}
# Random Forest: the most correlated features
start.time <- Sys.time()
model.corr.features <- randomForest(
  x = training.data[, most.correlated.outcome.feature.indices], 
  y = training.data$classe,
  xtest = cv.data[, most.correlated.outcome.feature.indices], 
  ytest = cv.data$classe, 
  ntree = number.trees,
  keep.forest = TRUE,
  proximity = TRUE,
  do.trace = FALSE)
end.time <- Sys.time()

model.summary <- rbind(model.summary, cbind( 
  trees = model.corr.features$ntree,
  features = ncol(training.data[, most.correlated.outcome.feature.indices]),
  model = "Correlated",
  training.accuracy = round((1 - sum(model.corr.features$confusion[,"class.error"])) * 100, digits = 3),
  cv.accuracy = round((1 - sum(model.corr.features$test$confusion[,"class.error"])) * 100, digits = 3),
  training.time = round(as.numeric(end.time - start.time, units = "secs"))
))
```

### Random Forest: the most correlated features with PCA

```{r}
# Apply PCA
pre.corr.pca <- preProcess(training.data[, most.correlated.outcome.feature.indices], method = "pca", thresh = 0.99)
training.data.corr.pca <- predict(pre.corr.pca, training.data[, most.correlated.outcome.feature.indices])
cv.data.corr.pca <- predict(pre.corr.pca, cv.data[, most.correlated.outcome.feature.indices])
testing.data.corr.pca <- predict(pre.corr.pca, testing.data[, most.correlated.outcome.feature.indices])

# Random Forest: the most correlated features with PCA
start.time <- Sys.time()
model.corr.pca.features <- randomForest(
  x = training.data.corr.pca, 
  y = training.data$classe,
  xtest = cv.data.corr.pca, 
  ytest = cv.data$classe, 
  ntree = number.trees,
  keep.forest = TRUE,
  proximity = TRUE,
  do.trace = FALSE)
end.time <- Sys.time()

model.summary <- rbind(model.summary, cbind( 
  trees = model.corr.pca.features$ntree,
  features = ncol(training.data.corr.pca),
  model = "Correlated + PCA",
  training.accuracy = round((1 - sum(model.corr.pca.features$confusion[,"class.error"])) * 100, digits = 3),
  cv.accuracy = round((1 - sum(model.corr.pca.features$test$confusion[,"class.error"])) * 100, digits = 3),
  training.time = round(as.numeric(end.time - start.time, units = "secs"))
))
```

### Random Forest: reduced features

```{r}
# Random Forest: reduced features
start.time <- Sys.time()
model.reduced.features <- randomForest(
  x = training.data[, -c(duplicated.feature.indices, label.index)], 
  y = training.data$classe,
  xtest = cv.data[, -c(duplicated.feature.indices, label.index)], 
  ytest = cv.data$classe, 
  ntree = number.trees,
  keep.forest = TRUE,
  proximity = TRUE,
  do.trace = FALSE)
end.time <- Sys.time()

model.summary <- rbind(model.summary, cbind( 
  trees = model.reduced.features$ntree,
  features = ncol(training.data[, -c(duplicated.feature.indices, label.index)]),
  model = "Reduced",
  training.accuracy = round((1 - sum(model.reduced.features$confusion[,"class.error"])) * 100, digits = 3),
  cv.accuracy = round((1 - sum(model.reduced.features$test$confusion[,"class.error"])) * 100, digits = 3),
  training.time = round(as.numeric(end.time - start.time, units = "secs"))
))
```

### Random Forest: reduced features with PCA

```{r}
# Apply PCA
pre.reduced.pca <- preProcess(training.data[, -c(duplicated.feature.indices, label.index)], method = "pca", thresh = 0.99)
training.data.reduced.pca <- predict(pre.reduced.pca, training.data[, -c(duplicated.feature.indices, label.index)])
cv.data.reduced.pca <- predict(pre.reduced.pca, cv.data[, -c(duplicated.feature.indices, label.index)])
testing.data.reduced.pca <- predict(pre.reduced.pca, testing.data[, -c(duplicated.feature.indices, label.index)])

# Random Forest: reduced features with PCA
start.time <- Sys.time()
model.reduced.pca.features <- randomForest(
  x = training.data.reduced.pca, 
  y = training.data$classe,
  xtest = cv.data.reduced.pca, 
  ytest = cv.data$classe, 
  ntree = number.trees,
  keep.forest = TRUE,
  proximity = TRUE,
  do.trace = FALSE)
end.time <- Sys.time()

model.summary <- rbind(model.summary, cbind( 
  trees = model.reduced.pca.features$ntree,
  features = ncol(training.data.reduced.pca),
  model = "Reduced + PCA",
  training.accuracy = round((1 - sum(model.reduced.pca.features$confusion[,"class.error"])) * 100, digits = 3),
  cv.accuracy = round((1 - sum(model.reduced.pca.features$test$confusion[,"class.error"])) * 100, digits = 3),
  training.time = round(as.numeric(end.time - start.time, units = "secs"))
))
```

### Summary of models

Summarizing the trained models.
```{r}
model.summary$trees = as.numeric(as.character(model.summary$trees))
model.summary$features = as.numeric(as.character(model.summary$features))
model.summary$training.accuracy = as.numeric(as.character(model.summary$training.accuracy))
model.summary$cv.accuracy = as.numeric(as.character(model.summary$cv.accuracy))
model.summary$training.time = as.numeric(as.character(model.summary$training.time))
model.summary
```

Plotting the cross-validating accuracy.  
```{r}
# Cross-validating accuracy chart
gg <- ggbarplot(
  model.summary, x = "model", y = "cv.accuracy",
  title = "Cross-validating accuracy (%)",      # Main title
  fill = "model",                               # Color by groups
  sort.val = "desc",                            # Sort value in descending order
  label = model.summary$cv.accuracy,            # Label values
  lab.pos = "in",                               # Position for labels
  lab.col = "white",                            # Color for labels
  lab.nb.digits = 2,                            # Number of decimal places (round)
  xlab = FALSE,                                 # X-axis labels
  ylab = FALSE,                                 # Y-axis labels
  rotate = FALSE,                               # Rotate vertically
  font.label = list(color = "white", size = 11, vjust = 1.5),
  ggtheme = theme_pubr()                        # ggplot2 theme
) 
ggpar(gg, legend = "none", legend.title = "")   # Customize legend
```

Plotting the training times in seconds.  
```{r}
# Training time chart
gg <- ggdotchart(
  model.summary, x = "model", y = "training.time",
  title = "Training time (sec)",                # Main title
  color = "model",                              # Color by groups
  sorting = "descending",                       # Sort value in descending order
  rotate = FALSE,                               # Rotate vertically
  label = model.summary$training.time,          # Label values
  add = "segments",                             # Add lolipop lines
  xlab = FALSE,                                 # X-axis labels
  ylab = FALSE,                                 # Y-axis labels
  dot.size = 10,                                # Large dot size
  y.text.col = TRUE,                            # Color y text by groups
  font.label = list(color = "white", size = 11, face = "bold", vjust = 0.5),
  ggtheme = theme_pubr(),                        # ggplot2 theme
) 
ggpar(gg, legend = "none", legend.title = "",  ylim = c(min(model.summary$training.time) * 0.95, max(model.summary$training.time) * 1.05)) + theme_cleveland()
```

## Testing models

Using all of trained models to make predictions on the test dataset (20 observations). The ground truth dataset consist of the expected good values.
```{r}
# Predictions of the models
test.predictions <- data.frame(row.names = 1:20, stringsAsFactors = TRUE)
test.predictions$ground.truth <- as.factor(c("B", "A", "B", "A", "A", "E", "D", "B", "A", "A", "B", "C", "B", "A", "E", "E", "A", "B", "B", "B"))
test.predictions$all <- predict(model.all.features, testing.data[,-label.index])
test.predictions$all.with.pca <- predict(model.pca.features, testing.data.pca)
test.predictions$correlated <- predict(model.corr.features, testing.data[, most.correlated.outcome.feature.indices])
test.predictions$corr.with.pca <- predict(model.corr.pca.features, testing.data.corr.pca)
test.predictions$reduced <- predict(model.reduced.features, testing.data[, -c(duplicated.feature.indices, label.index)])
test.predictions$reduced.with.pca <- predict(model.reduced.pca.features, testing.data.reduced.pca)
test.predictions

# Calculating testing accuracies
model.summary$testing.accuracy <- c(
  sum(equals(test.predictions$ground.truth, test.predictions$all)) / nrow(test.predictions) * 100,
  sum(equals(test.predictions$ground.truth, test.predictions$all.with.pca)) / nrow(test.predictions) * 100,
  sum(equals(test.predictions$ground.truth, test.predictions$all.with.pca)) / nrow(test.predictions) * 100,
  sum(equals(test.predictions$ground.truth, test.predictions$correlated)) / nrow(test.predictions) * 100,
  sum(equals(test.predictions$ground.truth, test.predictions$corr.with.pca)) / nrow(test.predictions) * 100,
  sum(equals(test.predictions$ground.truth, test.predictions$reduced.with.pca)) / nrow(test.predictions) * 100
)
```

Plotting the final testing accuracy for 20 observations.
```{r}
# Testing accuracy chart
gg <- ggbarplot(
  model.summary, x = "model", y = "testing.accuracy",
  title = "Testing accuracy (%)",                # Main title
  fill = "model",                                # Color by groups
  sort.val = "desc",                             # Sort value in descending order
  label = model.summary$testing.accuracy,        # Label values
  lab.pos = "in",                                # Position for labels
  lab.col = "white",                             # Color for labels
  lab.nb.digits = 2,                             # Number of decimal places (round)
  xlab = FALSE,                                  # X-axis labels
  ylab = FALSE,                                  # Y-axis labels
  rotate = FALSE,                                # Rotate vertically
  font.label = list(color = "white", size = 11, vjust = 1.5),
  ggtheme = theme_pubr()                         # ggplot2 theme
) 
ggpar(gg, legend = "none", legend.title = "")    # Customize legend
```

## Conclusion

Using all features to train the model give us an appropriate cross-validating and testing accuracy.  
However the reduced features provide more validating accuracy.

Plotting the error rate tendency of the first trained model.  
```{r}
plot(model.all.features, main = "Error rate tendency")
```
The error rate doesn't decline a lot after 50 trees.
